---
title: "Normalize Seurat Object"
author: "TLusardi"
date: "9/27/2019"
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    toc: yes
    toc_float: yes
params:
  run2process: "wk36_mpssr_wt"
  sourceDir: "home/groups/CEDAR/braun/seurat_obj"
  destDir: "/home/groups/CEDAR/braun/seurta_obj"
  file2save: "default"
  local: FALSE
  saveObj: TRUE
---

1. Read Seurat Object
2. Normalize/Scale using SCTransform
3. Create plots
4. Write updated object

Consider adding an evaluation of SCTransform vs. Normalize/Scale
  + Compare the methods - consider n-features

```{r setup_libs, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(sctransform)
library(data.table)
library(knitr)
library(ggplot2)
library(grid)
library(gridExtra)

run2process = params$run2process
saveObj = params$saveObj
```

## Read Seurat Object - `r run2process`

```{r setup_vars, include=FALSE}
# Could (should?) probably make these parameters that can be passed in
directory = list(rda = "/Users/lusardi/Documents/CEDAR/Projects/3.ASXL_mutant/analysis/data/rda")
```
Source Data Directory:  `r directory$raw`   
Processed Data Directory:  `r directory$rda`   
Processed Run:  `r run2process`   

```{r, readObject, echo=FALSE}
files2choose = list.files(directory$rda)
files2choose <- files2choose[!grepl("scTxform", files2choose)]
filename = sprintf("%s_SeuratObj", run2process)
# If there are multiple files with the object name, choose the last one in the list. This may not be the best option
selectmulti = sum(grepl(filename, files2choose))
file2open = files2choose[grepl(filename, files2choose)]
if (sum(grepl(filename, files2choose)) > 1) {message(sprintf("WARNING!!! %s had multiple options. Chose %s", run2process, file2open))}
seuratObj = readRDS(paste(directory$rda, file2open, sep = "/"))
Idents(seuratObj) <- "orig.ident"
```

### Calculate % Mitochondrial Genes per cell
```{r, calcMito, echo=FALSE}
seuratObj[["percent.mt"]] <- PercentageFeatureSet(object = seuratObj, pattern = "^mt-")

#Visualize QC metrics as a violin plot
VlnPlot(object = seuratObj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0.3)

# Plot density function of percent.mt
# Consider how much to include/exclude
colors = c('#e41a1c','#377eb8','#4daf4a','#984ea3','#ff7f00','#ffff33','#a65628','#f781bf','#999999')
probs = c(0.99, 0.95, 0.9, 0.85, 0.80)
cut_opts = quantile(seuratObj@meta.data$percent.mt, probs = probs)
p1 <- plot(density(seuratObj@meta.data$percent.mt), main = sprintf("Distribution of %% Mitochondrial Cells %s", "Wildtype"))
for (i in 1:length(cut_opts)) {
  p1 <- abline(v = cut_opts[i], col = colors[i])
}
legend_text = paste(probs, round(cut_opts, 1), sep = " - ")
p1 <- legend(x = "topright", legend = legend_text, col = colors[1:length(probs)], pch = 16)
```

***
### Calculate SCTransform

```{r, scTx, warning=FALSE, echo=FALSE}
print("Running SCT.all")
seuratObj.all <- SCTransform(seuratObj, verbose = FALSE)
print("Running SCT.regMT")
seuratObj.regMT <- SCTransform(seuratObj, vars.to.regress = "percent.mt", verbose = FALSE)
```

### Dimensionality reduction
```{r, dimReduce, warning=FALSE, echo=FALSE}
myassay = "SCT.all"
  print(sprintf("Dim reduct on %s", myassay))
  seuratObj.all <- RunPCA(seuratObj.all, verbose = TRUE) 
  seuratObj.all <- RunUMAP(seuratObj.all, dims = 1:30, verbose = FALSE)
  seuratObj.all <- FindNeighbors(seuratObj.all, dims = 1:30, verbose = FALSE)
  seuratObj.all <- FindClusters(seuratObj.all, verbose = FALSE)
  DimPlot(seuratObj.all, label = TRUE) + NoLegend()
  
myassay = "SCT.regMT"
  print(sprintf("Dim reduct on %s", myassay))
  seuratObj.regMT <- RunPCA(seuratObj.regMT, verbose = TRUE) 
  seuratObj.regMT <- RunUMAP(seuratObj.regMT, dims = 1:30, verbose = FALSE)
  seuratObj.regMT <- FindNeighbors(seuratObj.regMT, dims = 1:30, verbose = FALSE)
  seuratObj.regMT <- FindClusters(seuratObj.regMT, verbose = FALSE)
  DimPlot(seuratObj.regMT, label = TRUE) + NoLegend()
```

***
### Plot Key Features for each SCT run

```{r, plotFeatures, echo=FALSE}
featureList = list(erythroid = c("Hbb-bs", "Hba-a1"),
                   megakaryocytes = c("Itga2b", "Pbx1"),
                   basophils = c("Prss34", "Ms4a2", "Fcer1a"),
                   MEP = c("Pf4", "Mpl"),
                   myeloid = c("Cd33", "Mmp9"),
                   mast = c("Alox15", "Gata2"), 
                   bCells = c("Igha"),
                   neutrophil = c("Elane", "Ctsg", "Mpo"),
                   monocyte = c("Irf8", "Ctsg", "Mpo")) 
for (myfeatures in names(featureList)) {
  p1 = FeaturePlot(seuratObj.all, features = featureList[[myfeatures]], min.cutoff = "q9")
  grid.arrange(p1, top = sprintf("%s - %s", "seuratObj.all", myfeatures))
  
  p2 = FeaturePlot(seuratObj.regMT, features = featureList[[myfeatures]], min.cutoff = "q9")
  grid.arrange(p2, top = sprintf("%s - %s", "seuratObj.regMT", myfeatures))
}
```

***
### Tabulate # cells by cluster
```{r, tabulate, echo=FALSE}
kable(table(Idents(seuratObj.all), seuratObj.all$MULTI_ID_HTO))
kable(table(Idents(seuratObj.regMT), seuratObj.regMT$MULTI_ID_HTO))
```

***
### Write the updated Seurat object to a file

```{r, saveObject, echo=FALSE}
if (saveObj) {
  file2save = params$file2save
  file2save = ifelse(file2save == "default", run2process, file2save)
  
  filename = sprintf("%s_SeuratObj_scTxform_all_%s.Rdata", file2save, Sys.Date())
  print(sprintf("Saving %s to file %s at %s", run2process, filename, directory$rda))
  saveRDS(seuratObj.all, sprintf("%s/%s", directory$rda, filename))
  
  filename = sprintf("%s_SeuratObj_scTxform_regMT_%s.Rdata", file2save, Sys.Date())
  print(sprintf("Saving %s to file %s at %s", run2process, filename, directory$rda))
  saveRDS(seuratObj.regMT, sprintf("%s/%s", directory$rda, filename))
}
```

***
#### Session Information
```{r, echo=FALSE}
sessionInfo()
```
